\section{Evaluation and Current Validation Status}

\subsection{Methodological Update}

Earlier drafts of this paper reported finalized microbenchmark and latency numbers as if they were backed by a stable, reproducible benchmarking harness inside the repository. The March 2026 code snapshot does not support that claim. The current repository contains substantial implementation evidence and test suites, but it does not preserve a benchmark artifact sufficient to justify the previously stated performance tables as current empirical fact.

For that reason, this revised section evaluates the system using two evidence classes:
\begin{enumerate}
    \item repository-level implementation artifacts; and
    \item local test execution performed against the current codebase on 2026-03-13.
\end{enumerate}

\subsection{Implemented Subsystems}

\begin{table}[htbp]
\centering
\caption{Implemented Subsystems in the March 2026 Snapshot}
\label{tab:implemented_snapshot}
\small
\begin{tabularx}{\linewidth}{@{}l l X@{}}
\toprule
\textbf{Subsystem} & \textbf{Status} & \textbf{Evidence} \\
\midrule
Consensus-native tx model & Implemented & Native Elone transaction/input/output variants and validator paths exist in the consensus tree. \\
SDK signing and builders & Implemented & \texttt{sdk/elone} contains MuSig2 aggregation, signing contexts, and FUND/UPDATE/SETTLE/SPLICE builders. \\
Off-chain engine/store split & Implemented & \texttt{elone-core} and \texttt{elone-store} provide commands, ports, and SQLite persistence. \\
Node RPC submission/query layer & Implemented & Low-level RPC methods exist for serialized tx submission and channel-index queries. \\
Channel-factory lifecycle & Partial & FORK/MERGE/REBALANCE builders exist, but lifecycle and persistence remain unstable in current tests. \\
PTLC production execution & Partial & PTLC commitments and local state exist, but production adaptor-signature and redemption-signature paths remain incomplete. \\
PSTT workflow & Not implemented & The current repository does not contain a standalone PSTT protocol or sidecar workflow. \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Observed Local Test Results}

We executed the following local checks against the current repository state:

\begin{table}[htbp]
\centering
\caption{Observed Test Results on 2026-03-13}
\label{tab:test_results_current}
\footnotesize
\begin{tabularx}{\linewidth}{@{}X X X@{}}
\toprule
\textbf{Command / Suite} & \textbf{Observed Result} & \textbf{Interpretation} \\
\midrule
\texttt{cargo test -p tondi-elone-sdk --quiet} & 74 SDK tests passed; 3 failures remained in \texttt{channel\_manager\_lifecycle\_tests}. & Core SDK functionality is substantial, but the fork/subchannel lifecycle manager is not yet fully aligned with live consensus limits. \\
\midrule
\texttt{cargo test -p elone-core -p tondi-elone-sdk --quiet} & The \texttt{elone-core/tests/engine\_m2\_m3\_tests.rs} suite reported 4 failures: \texttt{test\_fork\_two\_subchannels\_merge\_both}, \texttt{test\_restart\_after\_splice\_subchannels\_survive}, \texttt{test\_splice\_history\_persisted}, and \texttt{test\_splice\_rebalance}. & The pure engine already supports splice planning/finalization, but channel-factory lifecycle and persistence behavior still require stabilization. \\
\midrule
Simnet / E2E Elone tests & Test files exist in the repository, but they were not re-run as part of this paper update. & The codebase contains broader integration intent, but those results should not be presented here as freshly validated evidence. \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{What the Current Evidence Supports}

The current evidence supports the following claims:
\begin{itemize}
    \item The core architectural thesis is implementable: Elone can be represented as a native, typed, dual-track UTXO protocol in a real Rust codebase.
    \item The safety-critical signing model is not hypothetical: the repository already contains typed sighash contexts, dual-sign UPDATE/SETTLE flows, and force-update / settle command paths.
    \item The system is not yet operationally complete: channel factories, PTLC execution, and cross-layer convergence still have meaningful engineering gaps.
    \item The current RPC surface is intentionally low-level and should not be confused with a finished wallet API.
\end{itemize}

\subsection{What the Current Evidence Does Not Yet Support}

The current repository snapshot does \textit{not} justify the following as settled empirical claims:
\begin{itemize}
    \item finalized microbenchmark speedup ratios relative to script-based systems;
    \item finalized end-to-end settlement latency distributions;
    \item production-readiness claims for channel-factory operation;
    \item claims that PSTT is already implemented and deployed.
\end{itemize}

Those statements may eventually become true, but they are not yet established by the present code snapshot.

\subsection{Revised Evaluation Summary}

The most accurate empirical summary of the March 2026 state is therefore:
\begin{enumerate}
    \item the implementation validates the protocol direction;
    \item the core consensus and SDK pieces are real and non-trivial;
    \item the current codebase remains pre-production;
    \item future quantitative evaluation should be rebuilt on top of a reproducible benchmark harness and refreshed integration runs.
\end{enumerate}
